---
title: "Introduction to helixvis"
author: "Raoul R. Wadhwa, Regina Stevens-Truss"
date: "25 September 2018"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to helixvis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: helixvis.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Prelude

The following code loads the helixvis package and the sample data that comes with it - these will be used later on in this vignette.

```{r loadpkg}
# load helixvis
library("helixvis")

# load dataset containing example peptides
data("sequence")
```

## Background

The helixvis R package allows users to construct 2-dimensional visualizations of 3-dimensional $\alpha$-helical oligopeptides in a programmatic, reproducible manner.
Currently, helixvis implements visualization of helical wheels, introduced by @wheeldiag, and wenxiang diagrams, introduced by @wenxiangdiag.

A prototypical $\alpha$-helix has approximately 3.6 residues per turn [@helixorig], implying a 100 degree rotation between consecutive residues.
After 18 rotations, the 19th residue returns to the same angle as the 1st residue.
Since helical wheels place all residues at an equal radius away from the center, having the same angle of rotation would cause complete overlap of the 1st and 19th residues.
As such, helical wheels are limited to plotting 18 residues in a single turn, although some tools do allow for plotting of more than 1 turn [@heliquest].
For consistency, we also limit wenxiang diagrams

## Helical wheels

Helical wheels provide a bird's eye view of an $\alpha$-helical peptides and are particularly helpful at highlighting prominent hydrophobic faces to explain a large hydrophobic moment [@eisenberg1982; @dathe1997] and design antimicrobial peptides [@chen2005; @jiang2011].

With helixvis, generating helical wheels takes a single line of code, as follows.

```{r wheel, fig.width = 4, fig.height = 4}
# print sequence that will be drawn
print(sequence[1, ])

# draw helical wheel
draw_wheel(sequence$Seq[1])
```

By default: nonpolar residues (e.g. glycine, tryptophan) will be colored grey; polar residues (e.g. serine, threonine) will be colored yellow; acidic residues with a positive charge (e.g. aspartic acid, glutamic acid) will be colored blue; and basic residues with a negative charge (e.g. lysine, arginine) will be colored red.
This color scheme can be modified using the `col` parameter, which accepts a character vector with 4 elements, each of which must be an element of `grDevices::colors()`.
The one-letter code for each residue can also be overlaid on each residue by setting the `labels` parameter to `TRUE`.

```{r wheelcol, fig.width = 4, fig.height = 4}
# draw helical wheel with custom color scheme
draw_wheel(sequence$Seq[1], col = c("pink", "orange", "white", "black"),
           labels = TRUE, label.col = "blue")
```

Sometimes, the dimensions of the graphics device being used are not squares, leading to awkward helical wheels (left panel of figure below).
Since helixvis uses the ggplot2 package to generate wheels, this can be corrected using `ggplot2::coord_fixed` (right panel of figure below).

```{r fixcoord, fig.width = 3, fig.height = 4, fig.show = "hold", fig.cap = "Left: awkward helical wheel due to dimensions of graphics device. Right: corrected helical wheel."}
# awkward looking wheel due to dimensions of graphical device
draw_wheel(sequence$Seq[1])

# corrected wheel using ggplot2::coord_fixed
draw_wheel(sequence$Seq[1]) +
  ggplot2::coord_fixed()
```

## Wenxiang diagrams

Wenxiang diagrams are quite similar to helical wheels, with the exception that distance from the first residue is represented as a larger radius.
Thus, wenxiang diagrams visually provide the order of amino acids in the sequence.
The code used to generate wenxiang diagrams is analogous to that for generation of helical wheels, with the sole difference being that the `draw_wenxiang` function is called instead of `draw_wheel`.

The following code blocks and figures reproduce the figures from the Helical Wheel section, redrawn as wenxiang diagrams.

```{r wenxiang, fig.width = 3, fig.height = 3}
# draw wenxiang diagram
draw_wenxiang(sequence$Seq[1])

# draw wenxiang diagram with custom color scheme
draw_wenxiang(sequence$Seq[1], col = c("pink", "orange", "white", "black"),
           labels = TRUE, label.col = "blue")
```

```{r wenxiangcoord, fig.width = 3, fig.height = 4, fig.show = "hold", fig.cap = "Left: awkward wenxiang diagram due to dimensions of graphics device. Right: corrected wenxiang diagram."}
# awkward looking wheel due to dimensions of graphical device
draw_wenxiang(sequence$Seq[1])

# corrected wheel using ggplot2::coord_fixed
draw_wenxiang(sequence$Seq[1]) +
  ggplot2::coord_fixed()
```

## Future directions

Current development of helixvis is focused on the following:

* allowing residue circle area to be proportional to biophysical measurements of amino acid size (surface area or volume)
* implement visualization of helical net diagrams, introduced by @netdiag.

## References